### 2013-12-27 paper/paper.tex
first full paper review

\begin{abstract}
Multi-objective optimization has been traditionally a matter of study in domains like engineering or finance, with little impact on games research. However, action-decision based on a multi-objective evaluation may be beneficial to obtain a high quality level of play. This paper presents a Multi-objective Monte Carlo Tree Search algorithm for planning and control in real-time game domains, those where the time budget to decide the next move to make is significantly reduced. A comparison is made between the proposed algorithm, a single-objective version of Monte Carlo Tree Search and a rolling horizon implementation of Non-dominated Sorting Evolutionary Algorithm II. Two different benchmarks are employed, the Deep Sea Treasure and the Multi-Objective Physical Travelling Salesman Problem. Using the same heuristics on each game, the analysis is focused on how well the algorithms explore the search space. The results described in this paper show that the proposed approach provides better results than the other algorithms it is compared with. Additionally, it is also shown that the algorithm is able to converge to different solutions of the optimal, or the best found during the search, Pareto front.
\end{abstract}

@@

Multi-objective optimization has been a field of study of manufacturing, engineering~\cite{Marler2004} and finance~\cite{Coello2006}, while having little impact in games research. This paper shows that multi-objective optimization has much to offer in developing game strategies that allow for a fine-grain control of alternative policies. The application of such approaches to this field can provide interesting results, especially in games that are long or complex enough where long-term planning is not trivial and achieving a good level of play requires balancing strategies.

@@

This paper proposes a multi-objective real-time algorithm version of Monte Carlo Tree Search (MCTS), a popular reinforcement learning approach within the last decade. The proposed algorithm is tested in two different games: a real-time version of the Deep Sea Treasure (DST; a classical multi-objective problem), and the Multi-Objective Physical Travelling Salesman Problem (MO-PTSP). The algorithm is also compared to two other approaches: a single-objective MCTS (that uses a weighted sum of features to value the state) and a rolling horizon Non-dominated Sorting Evolutionary Algorithm II (NSGA-II). This paper extends and formalizes our previous work described in~\cite{Perez2013a}.

Two main goals can be identified in this paper: first, the proposed algorithm must be applicable to real-time domains (those where the next move to make must be decided within a small time budget) and it should obtain better or at least the same performance than the other state of the art algorithms. It is important to highlight that all three algorithms tested employ the same heuristic functions to evaluate the features of a given game state. Thus, the focus of this research is set on how the algorithm explores the search space, instead of providing the best possible solution to each given problem.

@@

The balance between exploration and exploitation is achieved by setting the value of $C$. Higher values of $C$ weight more the second term of the UCB1 Equation~\ref{eq:ucb1}, giving preference to those actions that have been explored less, at the expense of taking actions with the highest average reward $Q(s,a)$. A commonly used value is $\sqrt{2}$, as it balances both facets of the search when the rewards are normalized between $0$ and $1$. It is worth noting that MCTS, when combined with UCB1 reaches asymptotically logarithmic regret~\cite{coquelin2007bandit}.

@@

A multi-objective optimization problem (MOOP) represents a scenario where two or more objective functions are to be optimized (either maximized or minimized). The general form of a MOOP is formally described as a maximization function $F_m(x)$, that transforms points in the decision space ($X$) to points in the solution space ($F$). The elements of the decision space are vectors of $n$ variables of the form $x = (x_1, x_2, \dots, x_n)$, while elements in the solution space are vectors with a dimension $m$: $F_m(x) = (f_1(x), f_2(x), \dots, f_m(x))$. Therefore, each solution provides $m$ different scores (or rewards, or fitness) that are meant to be optimized. Without loss of generality, it is assumed from now on that all objectives must be maximized.

It is said that a solution $F_m(x)$ \textit{dominates} another solution $F_m(y)$ if:

@@

However, there are some cases where it cannot be said that $F_m(x) \preceq F_m(y)$ or $F_m(y) \preceq F_m(x)$. This situation occurs when one of the objectives is better in $F_m(x)$ but a different objective is better in $F_m(y)$ (for instance, when $F_1(x) < F_1(y)$ but $F_2(x) > F_2(y)$). In this case, it is said that these solutions are non-dominated with respect to each other. Solutions that are not dominated by each other are grouped in a \textit{non-dominated set}. Given a non-dominated set $P$, it is said that $P$ is the \textit{optimal Pareto front} if there is no other solution in the solution space that dominates any member of $P$. The relation between decision and objective space, dominance and Pareto fronts is depicted in Figure~\ref{fig:moop}.

\begin{figure} [!t]
	\begin{center}
	\includegraphics[scale=0.235]{img/moopDec.png}
	\caption{Decision and Solution spaces in a MOOP with two variables ($x_1$ and $x_2$) and two objectives ($f_1$ and $f_2$). In the objective space, yellow dots are non-optimal objective vectors, while blue dots form the Pareto-optimal front.}
	\label{fig:moop}
	\end{center}
\end{figure}

As many Pareto front can be formed by several points in the solution space, it is important to devise a mechanism to assess the quality of a given front. A possibility is to use the Hypervolume Indicator (HV): given a Pareto front $P$, $HV(P)$ is defined as the volume of the objective space dominated by $P$. More formally, $HV(P) = \mu (x \in \mathbb{R}^{d} : \exists r \in P \hspace{0.25cm} s.t \hspace{0.25cm} r \preceq x)$, where $\mu$ is the de Lebesgue measure on $\mathbb{R}^{d}$. If the objectives are to be maximized, the higher the $HV(P)$, the better the front calculated. Figure~\ref{fig:hv} shows an example of $HV(P)$ where the objective dimension space is $2$.

@@

Many different algorithms have been proposed to tackle multi-objective optimization problems in the literature. One of the most widely known and used methods is the weighted-sum approach. The procedure consists of giving a weight to each one of the objective and produce a single result as the linear combination of objectives and weights. By varying the weights provided, it is possible to converge to different solutions of the optimal Pareto front, if this is convex. However, K. Deb~\cite{Deb2001} explains how linear scalarization approaches fail in those scenarios where the optimal Pareto front is non-convex.

@@

Multi-objective Reinforcement Learning (MORL)~\cite{Vamplew2010} changes this definition by using instead a vector $R = {r_0, r_1, \dots, r_n}$ as rewards of the problem. Thus, MORL problems differ from RL in having more than one objective objectives ($n$) that must be maximized. If the objectives are independent or they do not oppose each other, scalarization technique approaches, as described above, could be suitable to tackle the problem. Essentially, this would mean to use a conventional RL algorithm on a single objective obtained by a weighted-sum of the multiple rewards. However, this is not always the case, as it is usual that the objectives are in conflict and the policy ($\pi$) must balance among them.

@@

The second type of algorithms, \textit{multiple-policy}, target to approximate the optimal Pareto front of the problem. An example of this type of algorithm is the one given by L. Barrett~\cite{Barrett2008}, who propose the Convex Hull Iteration Algorithm. This algorithm provides the optimal policy for any linear preference function, by learning all policies that define the convex hull of the Pareto front.

@@

The next step is to define the value for each pair of state and action, $W(s,a)$, as in Equation~\ref{eq:wang}. $\overline{r}^p_{sa}$ is the projection of $\overline{r}_{sa}$ into the piecewise linear surface defined by the Pareto archive $P$ (see Figure~\ref{fig:wangHV}. Then, $HV(P \cup \overline{r}_{sa})$ is declared as the HV of $P$ plus the point $\overline{r}_{sa}$. If $\overline{r}_{sa}$ is dominated by P, the distance between $\overline{r}_{sa}$ and $\overline{r}^p_{sa}$ is subtracted from the HV calculation. The tree policy selects actions based on a maximization of the value of $W(s,a)$.

@@

The proposed algorithm was employed successfully in two domains: the DST and the Grid Scheduling problem, matching state of the art results in both domains, at the expense of a high computational cost.

@@

In the algorithm proposed in this paper, the vector $\overline{r}$ of rewards that was obtained at the end of an Monte Carlo simulation is back-propagated through the nodes visited in the last iteration until the root is reached. In the vanilla algorithm, each node would use this vector $\overline{r}$ to update its own accumulated reward vector $\overline{R}$. Instead of doing this, each node in the MO-MCTS algorithm keeps a local Pareto front ($P$), updated at each iteration with the reward vector $\overline{r}$ obtained at the end of the simulation. Algorithm~\ref{alg:updPareto} describes how the node statistics are updated.

@@

In this algorithm, if $\overline{r}$ is not dominated by the local Pareto front, it is added to the front and $\overline{r}$ is propagated to its parent. In case $\overline{r}$ is dominated by the node's Pareto front, the local Pareto front and there is no need to keep this propagation up the tree.

Three observations can be made about the mechanism described here:

\begin{itemize}
\item Each node in the tree has an estimate of the quality of the solutions reachable from there, both as an average (as in the baseline MCTS) and as the best case scenario (by keeping the non-dominated front $P$).
\item By construction, if a reward $\overline{r}$ is dominated by the local front of a node, it is a given that it will be dominated by the nodes above in the tree, so there is no need to update the fronts of the upper nodes, producing little cost in the efficiency of the algorithm.
\item It is easy to infer, from the last point, that the Pareto front of a node cannot be worse than the front of its children (in other words, the front of a child will never dominate that of its parent). Therefore, the root of the tree contains the best non-dominated front ever found during the search.
\end{itemize}

@@

This algorithm, same as NSGA-II due to their multi-objective nature, provide a non-dominated front as a solution. However, in planning and control scenarios like the games analyzed in this research, an action must be provided to perform a move in the next step. The question that arises is how to choose the move to make with the information available.

As shown before, it is straightforward to obtain which actions lead to what points in the front given as a solution: first gene in an NSGA-II individual, a root's child in MO-MCTS. Hence, by identifying the point in the Pareto front that the algorithm should converge to, 
it is possible to execute the action that leads to that point.

In order to select the point in the Pareto front, a weight vector $W (\forall w_i \in W, w_i \in [0,1])$ can be provided with the number of objectives as its dimension. Two different mechanisms are proposed, for reasons that will be explained in the experiments section (\ref{sec:exp}):

\begin{itemize}
\item Weighted Sum: the action chosen is the one that maximizes the weighted sum of the solution vector multiplied by $W$, for each point in the front.
\item Euclidean distance: the euclidean distance from each point in the Pareto front (normalized in $[0,1]$) to the vector $W$ is calculated. The action to choose would be the one that leads to the point in the Pareto front with the smaller distance to $W$. 
\end{itemize}

@@

Two different are used in this research to analyze the performance of the algorithm proposed.

@@

The Deep Sea Treasure (DST) is a well known multi-objective problem introduced by Vamplew et al.~\cite{Vamplew2010}. In this single-player puzzle, the agent moves a submarine with the objective of finding a treasure located at the bottom of the ship. The world is divided into a grid of $10$ rows and $11$ columns, and the vessel starts at the top left board position. There are three types of cells: empty cells (or water), that the submarine can traverse; ground cells that, as the edges of the grid, cannot be traversed; and treasure cells, that provide different rewards and finish the game. Figure~\ref{fig:dstBase} (on the left) shows the environment of the DST.

@@

The ship can perform four different moves: \textit{up}, \textit{down}, \textit{right}  and \textit{left}. If the action applied takes the ship off the grid or into the sea floor, the vessel's position does not change.  There are two objectives in this game: the number of moves performed by the ship, that must be minimized, and the value of the treasure found, that should be maximized. As can be seen in Figure~\ref{fig:dstBase}, the most valuable treasures are at a greater distance from the initial position, making these objectives orthogonal.

@@

Section~\ref{sec:moo} introduced the problems of linear scalarization approaches when facing non-convex optimal Pareto front. The concave shape of the DST's optimal front makes those approximations to converge to the non dominated solutions located at the edges of the Pareto front ($(-19,1)$,$(-1,124)$). Note that this happens independently from the weights chosen for the linear approximation: some solutions of the front just can't be reached with these approaches. Thus, successful approaches should be able to find all elements of the optimal Pareto front and converge to any of the non dominated solutions.

@@

In this example, the submarine starts in the initial position ($P1$, root of the tree) and makes a sequence of moves that places it in $P2$. From this location, if the vessel wishes to move to $P3$, two optimal trajectories would be route \textit{a} and \textit{b}. In a tree that does not use TT, there would be two different nodes to represent $P3$, although the location and number of moves performed up to this point are the same (thus, the states are equivalent). It is worthwhile to highlight that the coordinates of the vessel are not enough to identify two equivalent states, but also the number of moves is needed: imagine a third route from $P2$ to $P3$ with the moves: \textit{Up}, \textit{Right}, \textit{Right}, \textit{Down}, \textit{Down}. As the submarine performs $5$ moves, the states are not the same, and the node where the ship is in $P3$ now would be two levels deeper in the tree.

@@

The Multi-Objective Physical Travelling Salesman Problem (MO-PTSP) is a game employed in a competition held in the IEEE Conference on Computational Intelligence in Games (CIG) in 2013, a modification of the Physical Travelling Salesman Problem (PTSP), previously introduced by Perez et al.~\cite{PerezCEC2012}. The MO-PTSP is a real-time game where the agent navigates a ship and must visit $10$ waypoints scattered around the maze. All waypoints must be visited to consider a game as complete, and a game ticks timer is reset every time a waypoint is collected, finishing the game prematurely (and unsuccessfully) if it reaches $800$ game steps before visiting another waypoint.

@@

Regarding the third objective, the ship can suffer damage in two different ways: by colliding with obstacles and driving through lava. In the former case, the ship can collide with normal obstacles (that subtract $10$ units of damage) and specially damaging obstacles ($30$ units). In the latter, lava lakes are abundant in the MO-PTSP levels and, in contrast with normal surfaces, they deal one unit of damage for each time step the ship spends over this type of surface. All this subtractions are deduced from an initial counter of $5000$ points of damage.

Figure~\ref{fig:sampleMap} shows an example of an MO-PTSP map, as drawn by the framework. Waypoints yet not visited are painted as blue circles, while those already collected are depicted as empty circles. The green ellipses represent fuel canisters, normal surfaces are drawn in brown and lava lakes are printed as red-dotted yellow surfaces. Normal obstacles black, damaging obstacles are drawn in red. Blue obstacles are elastic walls that produce no damage to the ship. The vessel is drawn as a blue polygon and its trajectory is traced with a black line.

\begin{figure} [!t]
	\begin{center}
	\includegraphics[width=0.7\columnwidth]{img/moptspmap}
	\caption{Sample MO-PTSP map.}
	\label{fig:sampleMap}
	\end{center}
\end{figure}

@@

A macro-action of length $L$ is defined as a repetition of a given action during $L$ consecutive time steps. The main advantage of macro-actions is that the control algorithms can see further in the future, and then reduce the implication of open-endedness of real-time games (see Section~\ref{sec:mcts}). As this reduces the search space significantly, better algorithm performance is allowed. A consequence of using macro-actions is also that the algorithm can employ $L$ consecutive steps to plan the next move (the next macro-action) to make. Therefore, instead of spending $40$ milliseconds, as in MO-PTSP, to define the next action, the algorithm can employ $40 \times L$ milliseconds for this task, and hence perform a more extensive search.

@@

In order to evaluate a game state in MO-PTSP, three different measures or rewards are taken, $r_t, r_f, r_d$, one for each one of the objectives time, fuel and damage, respectively. All these rewards are defined so they have to be maximized. The first reward uses a measure of distance for the time objective, as indicated in Equation~\ref{eq:dist}:

@@

$d_s$ is the damage suffered so far in the game, and $d_m$ is the maximum damage the ship can take. In this case, three different variables are used to regulate the bahaviour of this objective: $\gamma$, $\beta_{1}$ and $\beta_{2}$. Both $\beta_{1}$ and $\beta_{2}$ have the same role as $\alpha$ in Equation~ref{eq:fuel}: they balance between the time objective and the damage measure. The difference is that $\beta_{1}$ is used in high speeds, while $\beta_{2}$ is employed with low velocities. This is distinguished by the parameter $\gamma$, that can be seen as a threshold value for the ship's speed ($s$). This differentiation is made in order to avoid low speeds in lava lakes, as this increase the damage suffered importantly. The values for these variables have been determined also empirically, and they are set to $\gamma = 0.8$, $\beta_{1} = 0.75$ and $\beta_{2} = 0.25$.

@@

The experiments performed in this research compare three different algorithms in the two benchmarks presented in Section~\ref{sec:bench}: a single objective MCTS (referred to here simply as \textit{MCTS}), the Multi-Objective MCTS (\textit{MO-MCTS}) and a rolling horizon version of the NSGA-II algorithm described in Section~\ref{sec:moo} (\textit{NSGA-II}). This NSGA-II version evolves a population where the individuals are sequences of actions (macro-actions in the MO-PTSP case), obtaining the fitness from the state of the game after applying such sequence.

All algorithms have a limited number of evaluations before providing an action to perform. In order for these games to be real-time, the time budget allowed is close to $40$ milliseconds. With the objective of avoiding congestion peaks at the machine where the experiments are run, the average of number of evaluations doable in $40$ ms. is calculated and employed in the tests. This leaded to $4500$ evaluations in the DST, and $500$ evaluations for MO-PTSP, using the same server where the PTSP and MO-PTSP competitions were run\footnote{Intel Core i5, 230GHz, 4GB of RAM}.

@@

As the optimal Pareto front of the DST is known, a measure of performance can be obtained by observing the percentage of times these solutions are found by the agents. As the solution that the algorithms converge to depends on the weights vector employed during the search (see end of Section~\ref{sec:momcts}), the approach taken here is to provide different weight vectors $W$ and analyze them separately.

@@

The results show clearly how approximating the optimal Pareto front allows for finding all possible solutions. Both NSGA-II and MO-MCTS approaches are able to converge to any solution in the front given the appropriate weight vector. It is important to highlight that these two algorithms employed the euclidean distance action selection (see Section~\ref{sec:momcts}). Experiments shown that weighted sum action selection provides similar results as in MCTS (this is, convergence to the edges of the front). The crucial distinction to make is that both algorithms NSGA-II and MO-MCTS allow for different action selection mechanisms, that are able to overcome this problem, by approximating a global Pareto front.

>>>>

\begin{abstract}
Multi-objective optimization has been traditionally a matter of study in domains like engineering or finance, with little impact on games research. However, action-decision based on a multi-objective evaluation may be beneficial to obtain a high quality level of play. This paper presents a Multi-objective Monte Carlo Tree Search algorithm for planning and control in real-time game domains, those where the time budget to decide the next move to make is significantly reduced. A comparison is made between the proposed algorithm, a single-objective version of Monte Carlo Tree Search and a rolling horizon implementation of Non-dominated Sorting Evolutionary Algorithm II. Two different benchmarks are employed, the Deep Sea Treasure and the Multi-Objective Physical Travelling Salesman Problem. Using the same heuristics on each game, the analysis is focused on how well the algorithms explore the search space. The results show that the proposed approach provides better results than the other algorithms it is compared with. Additionally, it is also shown that the algorithm is able to converge to different solutions of the optimal, or the best found during the search, Pareto front.
\end{abstract}

@@

Multi-objective optimization has been a field of study of manufacturing, engineering~\cite{Marler2004} and finance~\cite{Coello2006}, while having little impact in games research. This paper shows that multi-objective optimization has much to offer in developing game strategies that allow for a fine-grain control of alternative policies. The application of such approaches to this field can provide interesting results, especially in games that are long or complex enough where long-term planning is not trivial, and achieving a good level of play requires balancing strategies.

@@

This paper proposes a multi-objective real-time algorithm version of Monte Carlo Tree Search (MCTS), a popular reinforcement learning approach within the last decade. The proposed algorithm is tested in two different games: a real-time version of the Deep Sea Treasure (DST; a classical multi-objective problem), and the Multi-Objective Physical Travelling Salesman Problem (MO-PTSP). The algorithm is also compared with another two approaches: a single-objective MCTS (that uses a weighted sum of features to value the state) and a rolling horizon Non-dominated Sorting Evolutionary Algorithm II (NSGA-II). This paper extends and formalizes our previous work described in~\cite{Perez2013a}.

Two main goals can be identified in this paper: first, the proposed algorithm must be applicable to real-time domains (those where the next move to make must be decided within a small time budget) and it should obtain better or at least the same performance than the other state-of-the-art algorithms. It is important to highlight that all three algorithms tested employ the same heuristic functions to evaluate the features of a given game state. Thus, the focus of this research is set on how the algorithm explores the search space, instead of providing the best possible solution to each given problem.

@@

The balance between exploration and exploitation is achieved by setting the value of $C$. Higher values of $C$ weight more the second term of the UCB1 Equation~\ref{eq:ucb1}, giving preference to those actions that have been explored less, at the expense of taking actions with the highest average reward $Q(s,a)$. A commonly used value is $\sqrt{2}$, as it balances both facets of the search when the rewards are normalized between $0$ and $1$. It is worth noting that MCTS, when combined with UCB1, reaches asymptotically logarithmic regret~\cite{coquelin2007bandit}.

@@

A multi-objective optimization problem (MOOP) represents a scenario where two or more objective functions are to be optimized (either maximized or minimized). The general form of a MOOP is formally described as a maximization function $F_m(x)$, that transforms points in the decision space ($X$) to points in the solution space ($F$). The elements of the decision space are vectors of $n$ variables of the form $x = (x_1, x_2, \dots, x_n)$, while elements in the solution space are vectors of dimension $m$: $F_m(x) = (f_1(x), f_2(x), \dots, f_m(x))$. Therefore, each solution provides $m$ different scores (or rewards, or fitness) that are meant to be optimized. Without loss of generality, it is assumed from now on that all objectives must be maximized.

It is said that a solution $F_m(x)$ \textit{dominates} another solution $F_m(y)$ if and only if:

@@

However, there are some cases where it cannot be said that $F_m(x) \preceq F_m(y)$ or $F_m(y) \preceq F_m(x)$. This situation occurs when one of the objectives is better in $F_m(x)$ but a different objective is better in $F_m(y)$ (for instance, when $f_1(x) < f_1(y)$ but $f_2(x) > f_2(y)$). In this case, it is said that these solutions are non-dominated with respect to each other. Solutions that are not dominated by each other are grouped in a \textit{non-dominated set}. Given a non-dominated set $P$, it is said that $P$ is the \textit{optimal Pareto front} if there is no other solution in the solution space that dominates any member of $P$. The relation between decision and objective space, dominance and Pareto fronts is depicted in Figure~\ref{fig:moop}.

\begin{figure} [!t]
	\begin{center}
	\includegraphics[scale=0.235]{img/moopDec.png}
	\caption{Decision and solution spaces in a MOOP with two variables ($x_1$ and $x_2$) and two objectives ($f_1$ and $f_2$). In the objective space, yellow dots are non-optimal objective vectors, while blue dots form the Pareto-optimal front.}
	\label{fig:moop}
	\end{center}
\end{figure}

As many Pareto fronts can be formed by several points in the solution space, it is important to devise a mechanism to assess the quality of a given front. A possibility is to use the Hypervolume Indicator (HV): given a Pareto front $P$, $HV(P)$ is defined as the volume of the objective space dominated by $P$. More formally, $HV(P) = \mu (x \in \mathbb{R}^{d} : \exists r \in P \hspace{0.25cm} s.t \hspace{0.25cm} r \preceq x)$, where $\mu$ is the de Lebesgue measure on $\mathbb{R}^{d}$. If the objectives are to be maximized, the higher the $HV(P)$, the better the front calculated. Figure~\ref{fig:hv} shows an example of $HV(P)$ where the objective dimension space is $2$.

@@

Many different algorithms have been proposed to tackle multi-objective optimization problems in the literature. One of the most popular methods used is the weighted-sum approach. The procedure consists of giving a weight to each one of the objectives and produce a single result as the linear combination of objectives and weights. By varying the weights provided, it is possible to converge to different solutions of the optimal Pareto front, if this is convex. However, K. Deb~\cite{Deb2001} explains how linear scalarization approaches fail in those scenarios where the optimal Pareto front is non-convex.

@@

Multi-objective Reinforcement Learning (MORL)~\cite{Vamplew2010} changes this definition by using instead a vector $R = {r_0, r_1, \dots, r_n}$ as rewards of the problem. Thus, MORL problems differ from RL in having more than one objective ($n$) that must be maximized. If the objectives are independent or they do not oppose each other, scalarization technique approaches, as described above, could be suitable to tackle the problem. Essentially, this would mean to use a conventional RL algorithm on a single objective where the global reward is obtained from a weighted-sum of the multiple rewards. However, this is not always the case, as it is usual that the objectives are in conflict and the policy ($\pi$) must balance among them.

@@

The second type of algorithms, \textit{multiple-policy}, target to approximate the optimal Pareto front of the problem. The algorithm proposed in this paper would fit in this category. An example of this type of algorithm is the one given by L. Barrett~\cite{Barrett2008}, who propose the Convex Hull Iteration Algorithm. This algorithm provides the optimal policy for any linear preference function, by learning all policies that define the convex hull of the Pareto front.

@@

The next step is to define the value for each pair of state and action, $W(s,a)$, as in Equation~\ref{eq:wang}. $\overline{r}^p_{sa}$ is the projection of $\overline{r}_{sa}$ into the piecewise linear surface defined by the Pareto archive $P$ (see Figure~\ref{fig:wangHV}). Then, $HV(P \cup \overline{r}_{sa})$ is declared as the HV of $P$ plus the point $\overline{r}_{sa}$. If $\overline{r}_{sa}$ is dominated by P, the distance between $\overline{r}_{sa}$ and $\overline{r}^p_{sa}$ is subtracted from the HV calculation. The tree policy selects actions based on a maximization of the value of $W(s,a)$.

@@

The proposed algorithm was employed successfully in two domains: the DST and the Grid Scheduling problem, matching state-of-the-art results in both domains, at the expense of a high computational cost.

@@

In the algorithm proposed in this paper, the vector $\overline{r}$ of rewards that was obtained at the end of an Monte Carlo simulation is back-propagated through the nodes visited in the last iteration until the root is reached. In the vanilla algorithm, each node would use this vector $\overline{r}$ to update its own accumulated reward vector $\overline{R}$. In the algorithm proposed here, each node in the MO-MCTS algorithm also keeps a local Pareto front ($P$), updated at each iteration with the reward vector $\overline{r}$ obtained at the end of the simulation. Algorithm~\ref{alg:updPareto} describes how the node statistics are updated in MO-MCTS.

@@

Here, if $\overline{r}$ is not dominated by the local Pareto front, it is added to the front and $\overline{r}$ is propagated to its parent. In case $\overline{r}$ is dominated by the node's Pareto front, the local Pareto front does not change and there is no need to keep this propagation up the tree. Three observations can be made about the mechanism described here:

\begin{itemize}
\item Each node in the tree has an estimate of the quality of the solutions reachable from there, both as an average (as in the baseline MCTS) and as the best case scenario (by keeping the non-dominated front $P$).
\item By construction, if a reward $\overline{r}$ is dominated by the local front of a node, it is a given that it will be dominated by the nodes above in the tree, so there is no need to update the fronts of the upper nodes, producing little impact on the computational cost of the algorithm.
\item It is easy to infer, from the last point, that the Pareto front of a node cannot be worse than the front of its children (in other words, the front of a child will never dominate that of its parent). Therefore, the root of the tree contains the best non-dominated front ever found during the search.
\end{itemize}

@@

This algorithm, same as in NSGA-II due to their multi-objective nature, provide a non-dominated front as a solution. However, in planning and control scenarios like the games analyzed in this research, an action must be provided to perform a move in the next step. The question that arises is how to choose the move to make with the information available.

As shown before, it is straightforward to obtain which actions lead to what points in the front given as a solution: first gene in the best NSGA-II individual, a root's child in MO-MCTS. Hence, by identifying the point in the Pareto front that the algorithm should converge to, 
it is possible to execute the action that leads to that point.

In order to select a point in the Pareto front, a weight vector $W$ can be defined, with a dimension $n$ equal to the number of objectives ($W = (w_1, w_2, \dots, w_n); \sum_{i}^{n} w_i = 1$). Two different mechanisms are proposed, for reasons that will be explained in the experiments section (\ref{sec:exp}):

\begin{itemize}
\item Weighted Sum: the action chosen is the one that maximizes the weighted sum of the reward vector multiplied by $W$, for each point in the front.
\item Euclidean distance: the euclidean distance from each point in the Pareto front (normalized in $[0,1]$) to the vector $W$ is calculated. The action to choose would be the one that leads to the point in the Pareto front with the smaller distance to $W$. 
\end{itemize}

@@

Two different games are used in this research to analyze the performance of the algorithm proposed.

@@

The Deep Sea Treasure (DST) is a popular multi-objective problem introduced by Vamplew et al.~\cite{Vamplew2010}. In this single-player puzzle, the agent moves a submarine with the objective of finding a treasure located at the bottom of the ship. The world is divided into a grid of $10$ rows and $11$ columns, and the vessel starts at the top left board position. There are three types of cells: empty cells (or water), that the submarine can traverse; ground cells that, as the edges of the grid, cannot be traversed; and treasure cells, that provide different rewards and finish the game. Figure~\ref{fig:dstBase} shows the DST environment.

@@

The ship can perform four different moves: \textit{up}, \textit{down}, \textit{right}  and \textit{left}. If the action applied takes the ship off the grid or into the sea floor, the vessel's position does not change.  There are two objectives in this game: the number of moves performed by the ship, that must be minimized, and the value of the treasure found, that should be maximized. As can be seen in Figure~\ref{fig:dstBase}, the most valuable treasures are at a greater distance from the initial position, so the objectives are in conflict.

@@

Section~\ref{sec:moo} introduced the problems of linear scalarization approaches when facing non-convex optimal Pareto fronts. The concave shape of the DST's optimal front makes those approximations to converge to the non dominated solutions located at the edges of the Pareto front: $(81,124)$ and $(99,1)$. Note that this happens independently from the weights chosen for the linear approximation: some solutions of the front just can't be reached with these approaches. Thus, successful approaches should be able to find all elements of the optimal Pareto front and converge to any of the non dominated solutions.

@@

In this example, the submarine starts in the initial position ($P1$, root of the tree) and makes a sequence of moves that places it in $P2$. From this location, two optimal trajectories to move to $P3$ would be route \textit{a} and \textit{b}. In a tree that does not use TT, there would be two different nodes to represent $P3$, although the location and number of moves performed up to this point are the same (thus, the states are equivalent). It is worthwhile to highlight that the coordinates of the vessel are not enough to identify two equivalent states, but also the number of moves is needed: imagine a third route from $P2$ to $P3$ with the moves: \textit{Up}, \textit{Right}, \textit{Right}, \textit{Down}, \textit{Down}. As the submarine performs $5$ moves, the states are not the same, and the node where the ship is in $P3$ now would be two levels deeper in the tree.

@@

The Multi-Objective Physical Travelling Salesman Problem (MO-PTSP) is a game employed in a competition held in the IEEE Conference on Computational Intelligence in Games (CIG) in 2013, a modification of the Physical Travelling Salesman Problem (PTSP), previously introduced by Perez et al.~\cite{PerezCEC2012}. The MO-PTSP is a real-time game where the agent navigates a ship and must visit $10$ waypoints scattered around the maze. All waypoints must be visited to consider a game as complete, and a game-tick counter is reset every time a waypoint is collected, finishing the game prematurely (and unsuccessfully) if it reaches $800$ game steps before visiting another waypoint.

@@

Regarding the third objective, the ship can suffer damage in two different ways: by colliding with obstacles and driving through lava. In the former case, the ship can collide with normal obstacles (that subtract $10$ units of damage) and specially damaging obstacles ($30$ units). In the latter, lava lakes are abundant in MO-PTSP levels and, in contrast with normal surfaces, they deal one unit of damage for each time step the ship spends over this type of surface. All these subtractions are deduced from an initial counter of $5000$ points of damage.

Figure~\ref{fig:sampleMap} shows an example of an MO-PTSP map, as drawn by the framework. Waypoints yet not visited are painted as blue circles, while those already collected are depicted as empty circles. The green ellipses represent fuel canisters, normal surfaces are drawn in brown and lava lakes are printed as red-dotted yellow surfaces. Normal obstacles are black and damaging obstacles are drawn in red. Blue obstacles are elastic walls that produce no damage to the ship. The vessel is drawn as a blue polygon and its trajectory is traced with a black line.

\begin{figure} [!t]
	\begin{center}
	\includegraphics[width=0.8\columnwidth]{img/moptspmap}
	\caption{Sample MO-PTSP map.}
	\label{fig:sampleMap}
	\end{center}
\end{figure}

@@

A macro-action of length $L$ is defined as a repetition of a given action during $L$ consecutive time steps. The main advantage of macro-actions is that the control algorithms can see further in the future, and then reduce the implication of open-endedness of real-time games (see Section~\ref{sec:mcts}). As this reduces the search space significantly, a better algorithm performance is allowed. A consequence of using macro-actions is also that the algorithm can employ $L$ consecutive steps to plan the next move (the next macro-action) to make. Therefore, instead of spending $40$ milliseconds, as in MO-PTSP, to define the next action, the algorithm can employ $40 \times L$ milliseconds for this task, and hence perform a more extensive search.

@@

In order to evaluate a game state in MO-PTSP, three different measures or rewards are taken, $r_t, r_f, r_d$, one for each one of the objectives: time, fuel and damage, respectively. All these rewards are defined so they have to be maximized. The first reward uses a measure of distance for the time objective, as indicated in Equation~\ref{eq:dist}:

@@

$d_s$ is the damage suffered so far in the game, and $d_m$ is the maximum damage the ship can take. In this case, three different variables are used to regulate the behaviour of this objective: $\gamma$, $\beta_{1}$ and $\beta_{2}$. Both $\beta_{1}$ and $\beta_{2}$ have the same role as $\alpha$ in Equation~\ref{eq:fuel}: they balance between the time objective and the damage measure. The difference is that $\beta_{1}$ is used in high speeds, while $\beta_{2}$ is employed with low velocities. This is distinguished by the parameter $\gamma$, that can be seen as a threshold value for the ship's speed ($s$). This differentiation is made in order to avoid low speeds in lava lakes, as this increase the damage suffered importantly. The values for these variables have been determined also empirically, and they are set to $\gamma = 0.8$, $\beta_{1} = 0.75$ and $\beta_{2} = 0.25$.

It is important to highlight again that these three rewards, $r_t$, $r_f$ and $r_d$, are the same for the three algorithms tested in the experiments.

@@

The experiments performed in this research compare three different algorithms in the two benchmarks presented in Section~\ref{sec:bench}: a single objective MCTS (referred to here simply as \textit{MCTS}), the Multi-Objective MCTS (\textit{MO-MCTS}) and a rolling horizon version of the NSGA-II algorithm described in Section~\ref{sec:moo} (\textit{NSGA-II}). This NSGA-II version evolves a population where the individuals are sequences of actions (macro-actions in the MO-PTSP case), obtaining the fitness from the state of the game after applying such sequence. The populations sizes, determined empirically, were set to $20$ and $50$ individuals for DST and MO-PTSP respectively. The value of $C$ in Equation~\ref{eq:ucb1} and~\ref{eq:moucb} is set to $\sqrt{2}$.

All algorithms have a limited number of evaluations before providing an action to perform. In order for these games to be real-time, the time budget allowed is close to $40$ milliseconds. With the objective of avoiding congestion peaks at the machine where the experiments are run, the average of number of evaluations doable in $40$ ms. is calculated and employed in the tests. This leaded to $4500$ evaluations in the DST, and $500$ evaluations for MO-PTSP, using the same server where the PTSP and MO-PTSP competitions were run\footnote{Intel Core i5, 230GHz, 4GB of RAM.}.

@@

As the optimal Pareto front of the DST is known, a measure of performance can be obtained by observing the percentage of times these solutions are found by the players. As the solution that the algorithms converge to depends on the weights vector employed during the search (see end of Section~\ref{sec:momcts}), the approach taken here is to provide different weight vectors $W$ and analyze them separately.

@@

The results show clearly how approximating the optimal Pareto front allows for finding all possible solutions. Both NSGA-II and MO-MCTS approaches are able to converge to any solution in the front given the appropriate weight vector. It is important to highlight that these two algorithms employed the euclidean distance action selection (see Section~\ref{sec:momcts}). Other experiments, not included in this paper, showed that weighted sum action selection provides similar results as in MCTS (this is, convergence to the edges of the front). The crucial distinction to make here is that both algorithms NSGA-II and MO-MCTS allow for better action selection mechanisms, that are able to overcome this problem, by approximating a global Pareto front.
